Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2024-04-13 manuscript/manuscript.tex
Done edits for overall perf

\begin{figure}[h!]
\centering
\includegraphics[width=0.7\textwidth]{fig-methods.pdf}
\caption{Case study methods. VA records in Sierra Leone between 2019-2022 from the HEAL-SL study were filtered for physician agreed records, and pre-processed for COD assignment by GPT-3.5, GPT-4, InterVA-5 and InSilicoVA models. The output models were evaluated with CSMF Accuracy and PCCC overall, by age group, and by COD and five-year age ranges for each age group.}\label{fig-methods}
\end{figure}

@@

Of the 6939 records, GPT-4 had the highest PCCC followed by GPT-3.5, InSilicoVA, and InterVA-5 at 0.61, 0.58, 0.44, and 0.43 PCCC respectively (Table \ref{tab-perf-overall}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.03 PCCC. CSMF accuracies were above 0.7 for all models at 0.79, 0.74, 0.74, and 0.78 for GPT-4, GPT-3.5, InSilicoVA, and InterVA-5 respectively. Figure \ref{fig-perf-agegroup} shows the PCCC performance across age groups with reference to the CSMF Accuracy. GPT-4 had the highest PCCC at 0.65 and 0.62 for the adult and neonatal age groups, while GPT-3.5 had the highest PCCC at 0.57 for the child age group with GPT-4 slightly below it at 0.54. GPT-3.5 had the second highest PCCC at 0.56 fo the adult age group, while InSilicoVA had the second highest PCCC for the neonatal age group at 0.56. Both InterVA-5 and InSilicoVA had the lowest PCCC for the adult and child age groups at 0.41 and 0.45 PCCC respectively, while InterVA-5 and GPT-3.5 had the lowest PCCC for the neonatal age group at 0.49 and 0.51 respectively. The PCCC for models had larger differences between the minimum and maximum PCCC for the adult and child age groups at 0.3 and 0.28 respectively than the neonatal age group at 0.12. For reference, all records were also coded by GPT-3.5, GPT-4, InterVA-5, and InSilicoVA for comparison to physician agreed records. For physician agreed records, models had 0.05 to 0.1 higher PCCC values and 0.01 to 0.05 higher CSMF accuracy values (see Figure \ref{fig-perf-allvsagree} in Appendix \ref{app-perf-misc}).

@@

Across the adult age ranges, GPT-4 had the highest PCCC values at or above 0.65, followed by GPT-3.5, while InterVA-5 and InSilicoVA had the lowest PCCC values at or below 0.47. GPT-4 PCCC was at or above 0.7 for young to middle age ranges between 10 to 49 years, and below 0.7 for older age ranges between 50 to 69 years (Figure \ref{fig-perf-agerange-adult}). Each age range had between 135 (4\%) to 575 (15\%) records with the 65-69 year age range having the most records (n=575, 15\%), while other age ranges ranged from 4-9\% of all records. Figure \ref{fig-perf-cod-adult} shows the highest and lowest performing models by PCCC across 18 adult CODs. The highest PCCC values were GPT-4 between 0.53 to 1 for 15 of 18 CODs (GPT-3.5 within 0.05 for 7 of the 15 CODs), InSilicoVA for road and transport injuries at 0.84 (InterVA-5 close at 0.81), GPT-3.5 for ill-defined at 0.57,and InterVA-5 for ischemic heart disease at 0.61 (GPT-4 close at 0.57). The lowest PCCC values were InterVA-5 between 0 to 0.79 for 11 of 18 CODs (InSilicoVA within 0.05 for 3 of the 11 CODs), InSilicoVA between 0.01 to 0.41 for 6 of 18 CODs (InterVA-5 within 0.05 for other cardiovascular diseases and diarrhoeal diseases), and GPT-3.5 for ischemic heart disease at 0.49 (InSilicoVA within 0.05) and road and transport injuries at 0.59. Suicide and chronic respiratory diseases had the largest differences between minimum and maximum PCCC at 1 and 0.81 respectively. Maternal conditions, road and transport injuries, and tuberculosis had higher PCCC values above or at 0.67 (with the exception of GPT-3.5 for road and transportation injuries as an outlier), where there were small differences between minimum and maximum PCCC ranging from 0.12 to 0.19. Ischemic heart disease and unspecified infections had lower PCCC values below or at 0.6, where there were small differences between minimum and maximum PCCC ranging at 0.02 and 0.17 respectively. Each adult COD had between 3 (\textless 1\%) to 799 records (21\%) with malaria (n=799, 21\%), other noncommunicable diseases (n=414, 11\%), and unspecified infections (n=389, 10\%) making 42\% of all records, suicide having less than 1\% of records, and other CODs between 1-9\% of all records. GPT-4 had the highest PCCC for both male and female records at 0.72 and 0.7 respectively, while all other models had PCCC at or below 0.63 (see Figure \ref{fig-perf-sex-adult} in Appendix \ref{app-perf-misc}).

@@

\section{Physician Agreed Verbal Autopsy Records}\label{app-data}

>>>>

@@

Of the 6939 records, GPT-4 had the highest PCCC followed by GPT-3.5, InSilicoVA, and InterVA-5 at 0.61, 0.58, 0.44, and 0.43 PCCC respectively (Table \ref{tab-perf-overall}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.03 PCCC. CSMF accuracies were above 0.7 for all models at 0.79, 0.74, 0.74, and 0.78 for GPT-4, GPT-3.5, InSilicoVA, and InterVA-5 respectively. Figure \ref{fig-perf-agegroup} shows the PCCC performance across age groups with reference to the CSMF Accuracy. GPT-4 had the highest PCCC at 0.65 and 0.62 for the adult and neonatal age groups, while GPT-3.5 had the highest PCCC at 0.57 for the child age group with GPT-4 performing slightly worse at 0.54. GPT-3.5 had the second highest PCCC at 0.56 fo the adult age group, while InSilicoVA had the second highest PCCC for the neonatal age group at 0.56. InSilicoVA and InterVA-5 performed the worse for both adult and child age groups at PCCC scores below 0.5, while GPT-3.5 performed the worse for the neonatal age group with a PCCC of 0.42. The PCCC for models had larger differences between the minimum and maximum PCCC for the adult age group at 0.24, than the child and neonatal age groups at 0.12 and 0.20 respectively. For reference, all records, including non-physician agreed records, were also coded by GPT-3.5, GPT-4, InterVA-5, and InSilicoVA for comparison to physician agreed records. For physician agreed records, models had 0.05 to 0.09 higher PCCC values and 0.01 to 0.05 higher CSMF accuracy values than for all records (see Figure \ref{fig-perf-allvsagree} in Appendix \ref{app-perf-misc}).

@@

Across the adult age ranges, GPT-4 had the highest PCCC values at or above 0.59, followed by GPT-3.5, while InterVA-5 and InSilicoVA had the lowest PCCC values at or below 0.46. GPT-4 PCCC was between 0.61 and 0.72 for young to middle age ranges between 10 to 49 years, and at or below 0.62 for older age ranges between 50 to 69 years (Figure \ref{fig-perf-agerange-adult}). Each age range had between 135 (4\%) to 575 (15\%) records with the 65-69 year age range having the most records (n=575, 15\%), while other age ranges ranged from 4-9\% of all records. Figure \ref{fig-perf-cod-adult} shows the highest and lowest performing models by PCCC across 18 adult CODs. The highest PCCC values were GPT-4 between 0.36 to 0.99 for 12 of 18 CODs (GPT-3.5 within 0.05 for 7 of the 15 CODs), InSilicoVA for road and transport injuries at 0.84 (InterVA-5 close at 0.81), GPT-3.5 for ill-defined at 0.57,and InterVA-5 for ischemic heart disease at 0.61 (GPT-4 close at 0.57). The lowest PCCC values were InterVA-5 between 0 to 0.79 for 11 of 18 CODs (InSilicoVA within 0.05 for 3 of the 11 CODs), InSilicoVA between 0.01 to 0.41 for 6 of 18 CODs (InterVA-5 within 0.05 for other cardiovascular diseases and diarrhoeal diseases), and GPT-3.5 for ischemic heart disease at 0.49 (InSilicoVA within 0.05) and road and transport injuries at 0.59. Suicide and chronic respiratory diseases had the largest differences between minimum and maximum PCCC at 1 and 0.81 respectively. Maternal conditions, road and transport injuries, and tuberculosis had higher PCCC values above or at 0.67 (with the exception of GPT-3.5 for road and transportation injuries as an outlier), where there were small differences between minimum and maximum PCCC ranging from 0.12 to 0.19. Ischemic heart disease and unspecified infections had lower PCCC values below or at 0.6, where there were small differences between minimum and maximum PCCC ranging at 0.02 and 0.17 respectively. Each adult COD had between 3 (\textless 1\%) to 799 records (21\%) with malaria (n=799, 21\%), other noncommunicable diseases (n=414, 11\%), and unspecified infections (n=389, 10\%) making 42\% of all records, suicide having less than 1\% of records, and other CODs between 1-9\% of all records. GPT-4 had the highest PCCC for both male and female records at 0.72 and 0.7 respectively, while all other models had PCCC at or below 0.63 (see Figure \ref{fig-perf-sex-adult} in Appendix \ref{app-perf-misc}).

@@

\section{Case Study Methods}

\begin{figure}[h!]
\centering
\includegraphics[width=0.625\textwidth]{fig-methods.pdf}
\caption{Case study methods. VA records in Sierra Leone between 2019-2022 from the HEAL-SL study were filtered for physician agreed records, and pre-processed for COD assignment by GPT-3.5, GPT-4, InterVA-5 and InSilicoVA models. The output models were evaluated with CSMF Accuracy and PCCC overall, by age group, and by COD and five-year age ranges for each age group.}\label{fig-methods}
\end{figure}

\clearpage
\section{Physician Agreed Verbal Autopsy Records}\label{app-data}
